\setcounter{chapter}{-1}
\chapter[Doob-Gillespie Stochastic Simulation]{Simulating ODEs Stochastically with Doob-Gillespie Algorithms}
\label{chap:doob_gillespie}

Summary: How to simulate discrete populations using continuous models, how to test the quality of those simulations, and how to seek bifurcations.

\bigskip

Prerequisites: Calculus, Probability and Statistics, and programming experience. Experience in Ordinary Differential Equations aids comprehension but the chapter does not require it.

\bigskip

%Instructional Time: 30 to 40 hours.

\bigskip

Learning Objectives
\begin{itemize}
\item Identify shortfalls in using deterministic, continuous-state models like ODEs for probabilistic, discrete-state real-life phenomena.
\item Explain the re-interpretation of ODEs underlying the Doob-Gillespie approach.
\item Convert ODE models to Doob-Gillespie simulations.
\item Study probabilistic properties of models with statistics.
\item Explore parameter space to assess model sensitivity and model reliability.
\item Explain why ODE solutions differ from the stochastic mean.
\item Describe bifurcations, detect them, and account for them in sensitivity analysis.
\item Sample distributions by the inverse-CDF method.
\end{itemize}

\bigskip

Learning Outcomes
\begin{itemize}
\item Make stream plots of solution trajectories to non-linear ODEs.
\item Provide examples of discrete-state and deterministic real-life systems.
\item Provide examples of continuous-state and stochastic real-life systems.
\item Interpret an ODE in terms of state-change events, and determine the PMF (probability mass function) governing which event is next to occur.
\item Find equilibria of an ODE and analyze their stability.
\item Prove that exponentially distributed random variables are memoryless, and that the only continuous random variables which are memoryless are all exponentially distributed.
\item Scale or parameterize a system of ODEs so that the state space is on the scale of individuals.
\item Identify the relevant state-change events in a system of ODEs and their effects.
\item Parameterize ODE tank mixing models, ecological population models, epidemiological population models, and other models.
\item Convert ODE tank mixing models, ecological population models, epidemiological population models, and other models to Doob-Gillespie simulation farms.
\item Know stream plots fail for the SIR model, and identify when stream plots are not useful.
\item Capture distributional behavior using heat maps of stochastic trajectories.
\item Collect statistical data from simulation results and perform parameter space exploration, controlling for qualitative changes in behavior.
\item Use parameter space exploration to estimate and interpret univariate model sensitivity.
\item Use model sensitivity as a proxy to assess model reliability.
\item Identify bimodal or multimodal output with heatmaps and informally describe how to handle multimodal data.
\end{itemize}
